---
title: "Test 3: HIT Zwift session"
format: html
editor: visual
---

## Set environment and load packages

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r}


library(tidyverse)
library(patchwork)

theme_set(theme_bw())
```

## Test 3: HIT Zwift session, distal placement

### Load data

```{r}

setwd("C:/Users/simen/Desktop/training_analyses/data/battle_of_the_heart_rate_sensors")

test_3_hrm_pro_dist_plac <- read_csv("test_3_hrm_pro.csv")
test_3_polar_vs_dist_plac <- read_csv("test_3_polar_vs.csv")
test_3_forerunner_25_dist_plac <- read_csv("test_3_forerunner_255.csv")
```

### Clean data

Clean names and select relevant columns

```{r}

test_3_hrm_pro_dist_plac <- test_3_hrm_pro_dist_plac |> 
  janitor::clean_names() |> 
  select(record_timestamp_s, record_heart_rate_bpm)

# Polar VS

test_3_polar_vs_dist_plac <- test_3_polar_vs_dist_plac |>
  janitor::clean_names() |> 
  select(record_timestamp_s, record_heart_rate_bpm)

# Forerunner 255

test_3_forerunner_25_dist_plac <- test_3_forerunner_25_dist_plac |>
  janitor::clean_names() |> 
  select(record_timestamp_s, record_heart_rate_bpm)
```

Synchronize by the timestamp column

```{r}

test_3_hrm_vs_forerunner_dist_plac <- 
  left_join(
    test_3_hrm_pro_dist_plac, 
    test_3_polar_vs_dist_plac, 
    by = "record_timestamp_s"
    ) |> 
  left_join(
    test_3_forerunner_25_dist_plac, 
    by = "record_timestamp_s", 
    suffix = c("_hrm", "_forerunner")
  ) |> 
  rename(
    hrm_pro = record_heart_rate_bpm.x,
    polar_vs = record_heart_rate_bpm.y,
    forerunner_255 = record_heart_rate_bpm
  ) |> 
  mutate(
    elapsed_time = record_timestamp_s - min(record_timestamp_s)
  )
```

### Plots

Heart rate as a function of time for all three devices for the whole session

```{r}

test_3_hrm_vs_forerunner_dist_plac |> 
  mutate(
    elapsed_time = elapsed_time / 60,
    hrm_pro = hrm_pro / 195 * 100,
    polar_vs = polar_vs / 195 * 100,
    forerunner_255 = forerunner_255 / 195 * 100
  ) |>
  filter(
    elapsed_time < 51
  ) |> 
  ggplot() +
  geom_line(
    aes(x = elapsed_time, y = hrm_pro, color = "Garmin HRM-Pro"),
    alpha = 0.7
  ) +
  geom_line(
    aes(x = elapsed_time, y = polar_vs, color = "Polar Verity Sense"),
    alpha = 0.7
  ) +
  geom_line(
    aes(x = elapsed_time, y = forerunner_255, color = "Garmin Forerunner 255"),
    alpha = 0.7
  ) +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_color_manual(
    values = c("Garmin HRM-Pro" = "#E41A1C", 
               "Polar Verity Sense" = "#377EB8", 
               "Garmin Forerunner 255" = "#s"
    )
  ) +
  labs(
    x = "Time (minutes)",
    y = "Heart rate (% of maximum)",
    color = "Heart rate sensor",
    title = "Test 3: HIT indoor session on Zwift",
    subtitle = "45/15 x 10 x 3, progressing from heavy to severe"
  )

setwd("C:/Users/simen/Desktop/training_analyses/figures/battle_of_the_heart_rate_sensors")

ggsave(file = "test_3_all_devices_hr_comparison_full_session.jpg",
       dpi = 300,
       height = 4,
       width = 8)
```

Absolute difference to HRM-Pro as a function of time for the whole session

```{r}

test_3_hrm_vs_forerunner_dist_plac |> 
  mutate(
    elapsed_time = elapsed_time / 60,
    diff_vs = abs(hrm_pro - polar_vs),
    diff_255 = abs(hrm_pro - forerunner_255)
  ) |>
  filter(
    elapsed_time < 51
  ) |> 
  ggplot() +
  geom_line(
    aes(x = elapsed_time,
        y = diff_vs,
        color = "Polar Verity Sense"),
    alpha = 0.8
  ) +
  geom_line(
    aes(x = elapsed_time,
        y = diff_255,
        color = "Garmin Forerunner 255"),
    alpha = 0.8
  ) +
  labs(
    title = "Test 3: HIT indoor session on Zwift, absolute difference to HRM-Pro",
    subtitle = "45/15 x 10 x 3, progressing from heavy to severe",
    y = "Absolute difference in heart rate (BPM)",
    x = "Time (minutes)",
    color = "Heart rate sensor"
  ) +
  theme(
    panel.grid.minor = element_blank()
  ) +
   scale_color_manual(
     values = c("Polar Verity Sense" = "#377EB8", 
                "Garmin Forerunner 255" = "#4DAF4A"
    )
  )

setwd("C:/Users/simen/Desktop/training_analyses/figures/battle_of_the_heart_rate_sensors")

ggsave(file = "test_3_all_three_devices_hr_comparison__full_session_abs_diff.jpg",
       dpi = 300,
       height = 4,
       width = 8)
```

Taking a closer look on the second interval

```{r}

test_3_hrm_vs_forerunner_dist_plac |> 
  mutate(
    elapsed_time = elapsed_time / 60,
    hrm_pro = hrm_pro / 195 * 100,
    polar_vs = polar_vs / 195 * 100,
    forerunner_255 = forerunner_255 / 195 * 100
  ) |>
  filter(
    elapsed_time >= 24 & elapsed_time < 34
  ) |> 
  ggplot() +
  geom_line(
    aes(x = elapsed_time, y = hrm_pro, color = "Garmin HRM-Pro"),
    alpha = 0.7
  ) +
  geom_line(
    aes(x = elapsed_time, y = polar_vs, color = "Polar Verity Sense"),
    alpha = 0.7
  ) +
  geom_line(
    aes(x = elapsed_time, y = forerunner_255, color = "Garmin Forerunner 255"),
    alpha = 0.7
  ) +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_color_manual(
    values = c("Garmin HRM-Pro" = "#E41A1C", 
               "Polar Verity Sense" = "#377EB8", 
               "Garmin Forerunner 255" = "#4DAF4A"
    )
  ) +
  labs(
    x = "Time (minutes)",
    y = "Heart rate (% of maximum)",
    color = "Heart rate sensor",
    title = "Test 3: HIT indoor session on Zwift",
    subtitle = "Taking a closer look on the second interval"
  )

setwd("C:/Users/simen/Desktop/training_analyses/figures/battle_of_the_heart_rate_sensors")

ggsave(file = "test_3_all_devices_hr_comparison_interval_2.jpg",
       dpi = 300,
       height = 4,
       width = 8)
```

Taking a closer look on the third interval

```{r}

test_3_hrm_vs_forerunner_dist_plac |> 
  mutate(
    elapsed_time = elapsed_time / 60,
    hrm_pro = hrm_pro / 195 * 100,
    polar_vs = polar_vs / 195 * 100,
    forerunner_255 = forerunner_255 / 195 * 100
  ) |>
  filter(
    elapsed_time >= 36 & elapsed_time < 46
  ) |> 
  ggplot() +
  geom_line(
    aes(x = elapsed_time, y = hrm_pro, color = "Garmin HRM-Pro"),
    alpha = 0.7
  ) +
  geom_line(
    aes(x = elapsed_time, y = polar_vs, color = "Polar Verity Sense"),
    alpha = 0.7
  ) +
  geom_line(
    aes(x = elapsed_time, y = forerunner_255, color = "Garmin Forerunner 255"),
    alpha = 0.7
  ) +
  theme(
    panel.grid.minor = element_blank()
  ) +
  scale_color_manual(
    values = c("Garmin HRM-Pro" = "#E41A1C", 
               "Polar Verity Sense" = "#377EB8", 
               "Garmin Forerunner 255" = "#4DAF4A"
    )
  ) +
  labs(
    x = "Time (minutes)",
    y = "Heart rate (% of maximum)",
    color = "Heart rate sensor",
    title = "Test 3: HIT indoor session on Zwift",
    subtitle = "Taking a closer look on the third interval"
  )

setwd("C:/Users/simen/Desktop/training_analyses/figures/battle_of_the_heart_rate_sensors")

ggsave(file = "test_3_all_devices_hr_comparison_interval_3.jpg",
       dpi = 300,
       height = 4,
       width = 8)
```
